The Castro AMR Simulation Code: Current and Future Developments

### 2020-01-28 astronum_2019_Castro/astronum-castro-2019.tex
some work on addressing referees

\newcommand{\isot}[2]{$^{#2}\mathrm{#1}$}
\newcommand{\avg}[1]{{\left \langle #1 \right \rangle}}
\newcommand{\Rbs}[1]{{\bf R} \left ( #1 \right )}
\newcommand{\Ic}{\bm{\mathcal{I}}}

@@

The SDC algorithm used in \castro\ follows the ideas of
\cite{dutt:2000,minion:2003}, and uses low order explicit advection
and implicit reaction updates in a correction equation that when
applied iteratively achieves high-order time-accuracy.  In an operator
split implementation, reaction and burning proceed without knowing
about the other process.  In contrast, the update in the SDC
formulation explicitly couples the processes together.  In \castro\ we
consider reactions and hydrodynamics, and write our conservative
system as:
\begin{equation}
\Uc_t = \Advs{\Uc} + \Rbs{\Uc}
\end{equation}
where $\Uc$ is the vector of conserved quantities, $\Advs{\Uc}$ is the
advective term (the divergence of the hydrodynamic fluxes along with
hydrodynamic sources), and $\Rbs{\Uc}$ are the reactive source terms.
The update is done at specific temporal nodes, the number of locations
of which are picked to give the desired temporal accuracy.  The update
from one time node $m$ to $m+1$ appears as:
    \begin{align}
      \label{eq:sdc:general}
      \avg{\Uc}^{{m+1},(k+1)} = \avg{\Uc}^{m,(k+1)}
            &+ \delta t_m \left [ \avg{\Advs{\Uc}}^{m,(k+1)} - \avg{\Advs{\Uc}}^{m,(k)} \right ] \nonumber \\
            &+ \delta t_m \left [ \avg{\Rbs{\Uc}}^{{m+1},(k+1)} - \avg{\Rbs{\Uc}}^{{m+1},(k)} \right ] \nonumber \\
            &+ \int_{t^m}^{t^{m+1}} dt  \left (\avg{\Advs{\Uc}}^{(k)} + \avg{\Rbs{\Uc}}^{(k)}\right)
   \end{align}
We use a finite-volume formalism, so $\avg{\Uc}$ is the spatial average of the
conserved state in a zone, $\avg{\Advs{\Uc}}^m$ is the average
advective term in a zone at time $t^m$, and $\avg{\Rbs{\Uc}}^{m+1}$
the average of the reactive source at time $t^{m+1}$.  The second
superscript on each term, $(k)$ or $(k+1)$, is the iteration counter.  This
update is an implicit equation for the new state,
$\avg{\Uc}^{{m+1},(k+1)}$.  The last term in the update is an integral
over the sources constructed from the previous iteration's values at
each time node.  Each iteration of the SDC method increases the
temporal order of accuracy by one, up to the order of accuracy with
which the integral is constructed.

@@

Both the new SDC method and GPU offloading are important to our
science goals, in particular the problem of modeling a flame spreading
across the neutron star during an X-ray burst (XRB).  We discussed the
computational challenges of modeling these events in
\cite{astronum:2018}.  In short: the largest scale we need to model is
the size of the neutron star, or at least several times the scale at
which rotation and the lateral pressure gradient balance (the Rossby
length).  The smallest scale we need to model is dictated by
accurately capturing the burning---we are interested in both the
energy release and nucleosynthesis.  Our first set of two-dimensional simulations
(\cite{astronum:2018}, Eiden et al.\ in preparation) used two physical approximations to
make the problem more tractable.  First, we used a higher rotation
rate than expected in order to reduce the Rossby length, allowing us
to model a smaller region of that star.  Second, we artificially
boosted the speed of the flame to reduce the duration we need to
model.  Our next set of calculations will relax these approximations.

>>>>

\newcommand{\Ub}{\mathbf{U}}
\newcommand{\Fb}{\mathbf{F}}

\newcommand{\ddx}[1]{{\frac{{\partial#1}}{\partial x}}}
\newcommand{\ddy}[1]{{\frac{{\partial#1}}{\partial y}}}
\newcommand{\ddz}[1]{{\frac{{\partial#1}}{\partial z}}}
\newcommand{\omegadot}{\dot{\omega}}
\newcommand{\isot}[2]{$^{#2}\mathrm{#1}$}
\newcommand{\avg}[1]{{\left \langle #1 \right \rangle}}
\newcommand{\Rbs}[1]{{\bf R} \left ( #1 \right )}
\newcommand{\Ic}{\bm{\mathcal{I}}}

@@

The SDC algorithm used in \castro\ follows the ideas of
\cite{dutt:2000,minion:2003}, and uses low order explicit advection
and implicit reaction updates in a correction equation that when
applied iteratively achieves high-order time-accuracy.  In an operator
split implementation, reaction and burning proceed without knowing
about the other process.  In contrast, the update in the SDC
formulation explicitly couples the processes together.  In \castro\ we
consider reactions and hydrodynamics, and write our conservative
system as:
\begin{equation}
\Uc_t = \Advs{\Uc} + \Rbs{\Uc}
\end{equation}
where $\Uc$ is the vector of conserved quantities, $\Advs{\Uc}$ is the
advective term (the divergence of the hydrodynamic fluxes along with
hydrodynamic sources), and $\Rbs{\Uc}$ are the reactive source terms.
The conserved state has the form:
\begin{equation}
\Uc = (\rho, \rho X_k, \rho u, \rho v, \rho w, \rho E)^\intercal
\end{equation}
where $\rho$ is the mass density, $X_k$ are the mass fractions of the
nuclear species, $\Ub = (u, v, w)$ is the Cartesian velocity vector,
and $E$ is the total specific energy, related to the specific internal
energy, $e$, as $E = e + |\Ub|^2/2$.  The advective term takes the form:
\begin{equation}
\Advs{\Uc} = -\nabla \cdot \Fb (\Ub) + {\bf H}
\end{equation}
where $\Fb(\Ub)$ are the hydrodynamical fluxes:
\begin{equation}
\Fb \cdot \hat{\bf x} = \left (\begin{array}{c} \rho u \\ \rho X_k u \\ \rho u^2 + p \\ \rho v u \\ \rho w u \\ (\rho E + p) u  \end{array} \right )
\quad
\Fb \cdot \hat{\bf y} = \left (\begin{array}{c} \rho v \\ \rho X_k v \\ \rho u v \\ \rho v^2 + p \\ \rho w v \\ (\rho E + p) v  \end{array} \right )
\quad
\Fb \cdot \hat{\bf z} = \left (\begin{array}{c} \rho w \\ \rho X_k w \\ \rho u w \\ \rho v w \\ \rho w^2 + p \\ (\rho E + p) w  \end{array} \right )
\end{equation}
Here, $p$ is the pressure, which can be evaluated through our equation
of state, $p = p(\rho, e, X_k)$.  Note that we do not include
transport terms in this system, but it is straightforward to add in a
thermal diffusion flux.  ${\bf H}$ captures any hydrodynamical source
terms (e.g., gravity).  Finally the reactive sources take the form:
\begin{equation}
\Rbs{\Uc} = (0, \rho \omegadot_k, 0, 0, 0, \rho \dot{\epsilon})^\intercal
\end{equation}
where $\omegadot_k$ is the creation rate of species $k$ and $\dot{\epsilon}$
is the specific energy generation rate from reactions.

In the SDC formalism~\cite{dutt:2000,minion:2003}, we start by writing our system in integral form:
\begin{equation}
\Uc^{n+1} = \Uc^n + \int_{t^n}^{t^{n+1}} \left [\Advs{\Uc} + \Rbs{\Uc} \right ] dt
\end{equation}
We next introduce the idea of iteratively approximating the solution.
We'll use the superscript $k$ to denote the current iteration and
$k+1$ for the new iteration.  We rewrite the integral as:
\begin{align}
\Uc^{n+1,(k+1)} = \Uc^{n,(k+1)} +
   &\int_{t^n}^{t^{n+1}} \left \{ \left [ \Advs{\Uc}^{(k+1)} - \Advs{\Uc}^{(k)} \right ]
   + \left [ \Rbs{\Uc}^{(k+1)} - \Rbs{\Uc}^{(k)} \right ] \right \} dt + \nonumber \\
  & \int_{t^n}^{t^{n+1}} \left [\Advs{\Uc}^{(k)} + \Rbs{\Uc}^{(k)} \right ] dt
\end{align}
For the first integral, we approximate these terms using a first order
update.  For advection, we simply use the value at the starting time
(analogous to a forward Euler approximation) and for the reactions, we
treat them implicitly by using the value at the ending time (analogous
to a backward Euler approximation).  To gain temporal accuracy, we
further subdivide the time into $M+1$ nodes, and perform this update
over all time nodes for each iteration.  Denoting the time nodes with
the index $m$, the updated from one time node $m$ to $m+1$ appears as:
    \begin{align}
      \label{eq:sdc:general}
      \avg{\Uc}^{{m+1},(k+1)} = \avg{\Uc}^{m,(k+1)}
            &+ \delta t_m \left [ \avg{\Advs{\Uc}}^{m,(k+1)} - \avg{\Advs{\Uc}}^{m,(k)} \right ] \nonumber \\
            &+ \delta t_m \left [ \avg{\Rbs{\Uc}}^{{m+1},(k+1)} - \avg{\Rbs{\Uc}}^{{m+1},(k)} \right ] \nonumber \\
            &+ \int_{t^m}^{t^{m+1}} dt  \left (\avg{\Advs{\Uc}}^{(k)} + \avg{\Rbs{\Uc}}^{(k)}\right)
   \end{align}
where $\delta t_m = t^{m+1} - t^m$ is the time difference between nodes.
We use a finite-volume formalism, so $\avg{\Uc}$ is the spatial
average of the conserved state in a zone, $\avg{\Advs{\Uc}}^m$ is the
average advective term in a zone at time $t^m$, and
$\avg{\Rbs{\Uc}}^{m+1}$ the average of the reactive source at time
$t^{m+1}$.  The second superscript on each term, $(k)$ or $(k+1)$, is
the iteration counter.  This update is an implicit equation for the
new state, $\avg{\Uc}^{{m+1},(k+1)}$.  The last term in the update is
an integral over the sources constructed from the previous iteration's
values at each time node.  Each iteration of the SDC method increases
the temporal order of accuracy by one, up to the order of accuracy
with which the integral is constructed.

@@

Both the new SDC method and GPU offloading are important to our
science goals, in particular the problem of modeling a flame spreading
across the neutron star during an X-ray burst (XRB).  We discussed the
computational challenges of modeling these events in
\cite{astronum:2018}.  In short: the largest scale we need to model is
the size of the neutron star, or at least several times the scale at
which rotation and the lateral pressure gradient balance (the Rossby
length).  The smallest scale we need to model is dictated by
accurately capturing the burning---we are interested in both the
energy release and nucleosynthesis.  Our first set of two-dimensional simulations
(\cite{astronum:2018,eiden:2019} used two physical approximations to
make the problem more tractable.  First, we used a higher rotation
rate than expected in order to reduce the Rossby length, allowing us
to model a smaller region of that star.  Second, we artificially
boosted the speed of the flame to reduce the duration we need to
model.  Our next set of calculations will relax these approximations.
